
---
title: "Continuous and Discrete Random Variables"
author: "Rebecca C. Steorts"
output: 
     beamer_presentation:
      includes: 
          in_header: custom2.tex
font-size: 8px
---


Agenda
===


- Continuous Random Variables
- Discrete Random Variables 



Continuous Random Variables
===

A continuous random variable (r.v.) can take on an uncountably infinite number of values. 

Given a probability density function (pdf), f(y), allows us to compute

$$P(a \leq Y \leq b) = \int_{a}^{b} f(y) \; dy.$$

Properties of continuous random r.v.'s:

- $\int f(y) \; dy = 1.$
- For any value $y$, $$P(Y=y) = \int_{y}^{y} f(y) \; dy = 0 \implies $$
$$P(y < Y) = P(y \leq Y).$$

Discrete Random Variables
===

A discrete random variable has a countable number of possible values, where the associated probabilities are calculated for each possible value using a probabilitiy mass function (pmf). 

A pmf is a function that calculates $P(Y = y)$, given each variable's parameters.

Common Discrete distributions
===

- Bernoulli
- Geometric
- Negative Binomial 
- Hypergeometric 

Common Continuous distributions
===

- Exponenial
- Gamma
- Normal (Gaussian)
- Beta

Beta distribution
===
Given $a,b>0$, we write $\theta \sim \Beta(a,b)$ to mean that $\theta$ has pdf

$$
p(\theta) =\Beta(\theta|a,b) =\frac{1}{B(a,b)}\theta^{a-1}(1-\theta)^{b-1}\I(0<\theta<1),
$$

i.e., $p(\theta)\propto \theta^{a-1}(1-\theta)^{b-1}$ on the interval from $0$ to $1$.

- Here, 
$$B(a,b) = \frac{\Gamma(a)\Gamma(b)}{\Gamma(a+b)}$$.

- Parameters $a$, $b$ control the shape of the distribution. 

- This distribution models random behavior of percentages/proportions.

Beta distribution
===

```{r, echo = FALSE}
#define range
p = seq(0,1, length=100)

#plot several Beta distributions
plot(p, dbeta(p, 2, 10), ylab='density', type ='l', col='purple')
lines(p, dbeta(p, 2, 2), col='red') 
lines(p, dbeta(p, 5, 2), col='blue')

#add legend
legend(.7, 4, c('Beta(2, 10)','Beta(2, 2)','Beta(1,1)'),
       lty=c(1,1,1),col=c('purple', 'red', 'blue'))

```

Beta distribution example
===

Suppose that a college models probabilities of student accepting admission via the $\Beta(a,b)$ distribution, where a,b > 0  are fixed and known. 

What is the probability that a randomly selected student has prob of accepting an offer larger than 80 percent, where  a=4/3 and b=2.

```{r}
pbeta(0.8, shape1 = 4/3, shape2 = 2, lower.tail = FALSE)
```




