
---
title: "Module 4: The Multivariate Normal Distribution"
author: "Rebecca C. Steorts"
date: Hoff, Section 7.4
output: 
     beamer_presentation:
      includes: 
          in_header: custom2.tex
font-size: 8px
---



Agenda
===
- Motivational reading comprehension case study 
- Introduction/Review of vectors, matrices
- Population means/covariance matrices
- General multivariate notation
- Background on linear algebra (with practice exercises)
- Determinants, traces, quadratic forms

Agenda
===

- The multivariate normal distribution (MVN)
- Exercise with the MVN
- Case study on reading comprehension

What you should learn
===

- You will learn background on linear algebra
- You will learn how to model multivariate data, where we consider
an application to reading comprehension tests
- You will learn the notation for multivariate random variables
- You will learn about the multivariate density of the normal 


Goal
===

The goal of this module is to be able **to understand how to work with multivariate distributions**, such as the multivariate normal distribution. 

We also want to understand how univariate models that we have used in the past translate to the multivariate setting. 

Before we can delve in, we must review background on matrices, vectors, and **multivariate notation**. We also must review background on **linear algebra**. 

Example: Reading Comprehension
===


A sample of 22 children are given reading comprehension tests before and after receiving a particular instructional method.\footnote{This example follows Hoff (Section 7.4, p. 112).}

Each student $i$ will then have two scores, $Y_{i,1}$ and $Y_{i,2}$ denoting the pre- and post-instructional scores respectively. 

Denote each student’s pair of scores by the vector $\bm{Y}_i$
$$
\bm{Y}_{i} = \left( \begin{array}{c}
Y_{i,1}\\
Y_{i,2}\\
\end{array} \right) 
= \left( \begin{array}{c}
\text{score on first test}\\
\text{score on second test}\\
\end{array} \right)
$$
where $i=1,\ldots,n$ and $p=2.$

Example: Reading Comprehension
===

$$\bm{X}_{n \times p} = 
\left( \begin{array}{cccc}
x_{11} & \textcolor{red}{x_{12}} & \ldots&  x_{1p}\\
x_{21} & \textcolor{red}{x_{22}} & \ldots& x_{2p} \\
x_{31} & \textcolor{red}{x_{32}} & \ldots& x_{3p} \\
x_{i1} & \textcolor{red}{x_{i2}} & \ldots& x_{ip} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n1} & \textcolor{red}{x_{n2}} &\ldots& x_{np}
\end{array} \right).
$$

- A row of $\bm{X}_{n \times p}$ represents a covariate we might be interested in, such as age of a person.

- Denote $x_{i}$ $(p \times 1)$ as the $i$th \textcolor{red}{row vector} of the $X_{n \times p}$ matrix.

\[  x_{i}= \left( \begin{array}{c}
x_{i1}\\
\textcolor{red}{x_{i2}}\\
\vdots\\
x_{ip}
\end{array} \right) \]

Example: Reading Comprehension
===

We may be interested in the population mean $\bmu_{p \times 1}.$

$$
E[\bm{Y}] =: E[\bm{Y}_{i}] = \left( \begin{array}{c}
Y_{i,1}\\
Y_{i,2}\\
\end{array} \right) 
=  \left( \begin{array}{c}
\mu_1\\
\mu_2\\
\end{array} \right) 
= 
\bmu
$$

Example: Reading Comprehension
===
We also may be interested in the population covariance matrix, $\Sigma_{p \times p}.$

By definition: 
\begin{align}
\Sigma  
&= Cov(\bm{Y})\\
&=
\left( \begin{array}{cccc}
E[Y_1^2] - E[Y_1]^2 & E[Y_1Y_2] - E[Y_1]E[Y_2] \\
E[Y_1Y_2] - E[Y_1]E[Y_2] & E[Y_2^2] - E[Y_2]^2
\end{array} \right)\\
&=
\left( \begin{array}{cccc}
\sigma_1^2 & \sigma_{1,2} \\
\sigma_{1,2} & \sigma_2^2
\end{array} \right)
\end{align}

Remark: $Cov(Y_1) = Var(Y_1) = \sigma_1^2. \qquad Cov(Y_1, Y_2) = \sigma_{1,2}.$

How do we expand this beyond our reading comprehension example
===

We introduced our notation based upon a specific example to reading comprehension. 

How can we make this more general and applicable to general case studies and problems?

General Notation
===
Assume that $\bm{y}_{p \times 1} \sim (\mu_{p \times 1}, \Sigma_{p \times p}).$

$$\bm{y}_{p \times 1}= \left( \begin{array}{c}
y_{1}\\
{y_{2}}\\
\vdots\\
y_{p}
\end{array} \right).$$



$$\bmu_{p \times 1}= \left( \begin{array}{c}
\mu_1\\
\mu_2\\
\vdots\\
\mu_p
\end{array} \right)
$$
$$
\sig_{p \times p} = Cov(\bm{y}) =
\left( \begin{array}{cccc}
\sigma_1^2 & \sigma_{12} & \ldots&  \sigma_{1p}\\
\sigma_{21} & \sigma_2^2 & \ldots& \sigma_{2p}\\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{p1} & \sigma_{p2} &\ldots& \sigma_p^2
\end{array} \right).
$$


Background
===

Before proceeding, we need to review some basic concepts from linear algebra:

1. Basic properties of matrices
2. Useful lemmas for working with matrices

The determinant of a matrix 
===

Assume a matrix $A_{n \times n}$  is invertible. The 
$$\det(A) = a_{i1}A_{i1} + a_{i2}A_{i2} + \cdots + a_{in}A_{in},$$
where $A{ij}$ are the co-factors and are computed from $$A_{ij} = (-1)^{i+j}det(M_{ij}).$$ 
$M_{ij}$ is known as the minor matrix and is the matrix you get if you eliminate row i and column j from matrix $A.$ You must apply this technique recursively.
\vskip 1em 

\textbf{We only use this technique when doing such calculations by hand or in proof-based approaches.}

The determinant of a matrix 
===

- How on earth do I use the complicated formula on the pervious slide. 

\textbf{Easy: Use the \texttt{det} command in \texttt{R} when faced with an application.}

\vskip 1em 

- You will also see a determinant in the definition of the multivariate normal distribution. 

\textbf{Important point: It's just a function and we typically do not need to evalute it in this course!}


The trace of a matrix 
===

Assume a matrix $H_{p \times p}.$

$$\text{trace}(H) = \sum_i h_{ii},$$ where $h_{ii}$ are the diagonal elements of $H.$



The trace of a matrix 
===

$$
H =
\left( \begin{array}{cc}
1 & 0 \\
0 & 1 \\
\end{array} \right).
$$
What is $\text{tr}({H})?$ 

\vline

(Take 1 minute to complete this.)

Linear Algebra Tricks
===

Suppose that A is $n \times n$ matrix and suppose
that B is a $n \times n$ matrix. 

Lemma 1:  

$$tr(AB) = tr(BA)$$
Proof: Exercise. (You can find the proof at the end of the slides to check your work).




Linear Algebra Tricks
===

Lemma 2: 

Suppose $\bm{x_{p \times 1}}$ is a vector and $A$ is a $p \times p$ dimensional matrix. 

Then $\bm{x_{1 \times p}}^TA_{p \times p}\bm{x_{p \times 1}}$ is called a **quadratic form**. 

$$\bm{x}^TA\bm{x} = tr(\bm{x}^TA\bm{x}) = tr(\bm{x}\bm{x}^TA) = tr(A\bm{x}\bm{x}^T)$$

Proof: Exercise. 

Linear Algebra Tricks
===

Proof of Lemma 2:

\begin{align}
tr({\bm{x}^TA\bm{x}})  
&= \sum_i (x^tAx)_{ii} \\ 
& = (\bm{x}^T(A\bm{x})) \\
& = tr(A\bm{x}\bm{x}^T) \; (\text{by Lemma 1})
\end{align}

\begin{align}
tr({\bm{x}^TA\bm{x}})  
&= \sum_i (x^tAx)_{ii} \\ 
& = ((\bm{x}^TA)\bm{x}) \\
& = tr(\bm{x}\bm{x}^TA)\; (\text{by Lemma 1})
\end{align}




Notation
===
\begin{itemize}
\item MVN is generalization of univariate normal.
\item For the MVN, we write $\bm{y} \sim
\mathcal{MVN}(\bmu,\Sigma)$. 
\item The $(i,j)^{\text{th}}$
component of $\Sigma$ is the covariance between $Y_i$ and~$Y_j$ (so
the diagonal of $\Sigma$ gives the component variances).
\end{itemize}

Example: $Cov(Y_1, Y_2)$ is just one element of the matrix $\Sigma.$

Multivariate Normal
===
Just as the probability density of a scalar normal is
\begin{equation}
p(x) = {\left(2\pi\sigma^2\right)}^{-1/2}\exp{\left\{ -\frac{1}{2} \frac{(x-\mu)^2}{\sigma^2}\right\}},
\end{equation}
the probability density of the multivariate normal is
\begin{equation}
p(\bm{x}) = {\left(2\pi\right)}^{-p/2}(\det{\Sigma})^{-1/2} \exp{\left\{-\frac{1}{2} (\bm{x}-\bmu)^T\Sigma^{-1} (\bm{x} - \bmu)\right\}}.
\end{equation}
\textcolor{blue}{Univariate normal is special case of the multivariate normal with a one-dimensional mean ``vector'' and a one-by-one variance ``matrix.''}

Standard Multivariate Normal Distribution
===

Lemma 3. 

Consider $$Z_1, \ldots, Z_n \stackrel{iid}{\sim} N(0,1).$$
Show that $$Z_1,\ldots,Z_n \sim MVN(\textbf{0},I_{n \times n}).$$

Proof of Lemma 3
===
Proof:

\begin{align}
f_z(z) &= \prod_{i=1}^n (2\pi)^{-1/2} e^{-z_i^2/2}\\
& = (2\pi)^{-n/2} e^{\sum_i-z_i^2/2}\\
& = (2\pi)^{-n/2} e^{-z^Tz/2}.
\end{align}
The last line is follows since  $\sum_i-z_i^2 = -z^Tz.$

Thus, $Z_1,\ldots,Z_n \sim \text{MVN}(\textbf{0},I).$





Likelihood
===

\begin{align}
p(\bm{y} \mid \btheta, \Sigma) &= \prod_{i=1}^n
{\left(2\pi\right)}^{-p/2}\det{\Sigma}^{-1/2} \exp{\left\{-\frac{1}{2} (y_i-\btheta)^T\Sigma^{-1} (y_i - \btheta)\right\}}\\
\propto 
& \exp-\frac{1}{2} {\left \{ \sum_i y_i^T \Sigma^{-1} y_i -2 \sum_i \btheta^T \Sigma^{-1} y_i + 
\sum_i \btheta^T\Sigma^{-1} \btheta 
 \right \}}\\
 & \propto \exp-\frac{1}{2} {\left \{  -2 \btheta^T \Sigma^{-1} n\bar{y} + 
n \btheta^T\Sigma^{-1} \btheta 
 \right \}}\\
  & \propto \exp-\frac{1}{2} {\left \{  -2 \btheta^T b_1+ 
\btheta^T A_1 \btheta \right \}},
\end{align}
where $$b_1= \Sigma^{-1} n\bar{y}, \quad A_1 = n\Sigma^{-1}$$ and 
$$\bar{y} := (\frac{1}{n}\sum_i y_{i1} ,\ldots, \frac{1}{n} \sum_i y_{ip})^T.$$









Working with Multivariate Normal Distribution
===
The `R` package, `mvtnorm`, contains functions for evaluating and simulating from a multivariate normal density.

```{r}
library(ggplot2)
library(MASS)
library(mvtnorm)
library(car)
```

Simulating Data
===
Simulate a single multivariate normal random vector using the `rmvnorm` function.

```{r}
# Each row corresponds to a sample
# Here we have one sample (one row)
rmvnorm(n = 1, mean = rep(0, 2), sigma = diag(2))
```

Simulation Study and Investigation
===

```{r}
n <- 500  # Number of samples
mu <- c(5, 10, 15)  # Mean vector
Sigma <- matrix(c(4, 2, 1, 
                  2, 5, 2, 
                  1, 2, 3), 
                ncol = 3)  # Covariance matrix
data <- mvrnorm(n = n, mu = mu, Sigma = Sigma)
colnames(data) <- c("X1", "X2", "X3")
data <- as.data.frame(data)
```

Scatterplot Matrix
===
```{r}
scatterplotMatrix(~ X1 + X2 + X3, data = data, 
                  main = "Scatterplot Matrix of Multivariate Data")
```


Marginal Histogram
===
```{r, echo = F}
# Marginal histograms with density plots
par(mfrow = c(1, 3))  # Layout for plots
for (i in 1:3) {
  hist(data[[i]], breaks = 30, probability = TRUE, 
       main = paste("Histogram of", colnames(data)[i]),
       xlab = colnames(data)[i])
  lines(density(data[[i]]), col = "red", lwd = 2)
}
```

Q-Q Plots
===
```{r, echo = F}
# Q-Q plots to assess normality
par(mfrow = c(1, 3))
for (i in 1:3) {
  qqnorm(data[[i]], main = paste("Q-Q Plot of", colnames(data)[i]))
  qqline(data[[i]], col = "blue", lwd = 2)
}
```

Not sure
===
```{r, echo = F, message = FALSE}
# Combine visualizations with ggplot2
library(ggplot2)
pair_plot <- GGally::ggpairs(data, 
                             lower = list(continuous = "smooth"),
                             diag = list(continuous = "densityDiag"),
                             upper = list(continuous = "cor"))
print(pair_plot)
```

<!-- Evaluation -->
<!-- === -->
<!-- Evaluate the multivariate normal density at a single value using the `dmvnorm` function. -->

<!-- ```{r} -->
<!-- dmvnorm(rep(0, 2), mean = rep(0, 2), sigma = diag(2)) -->
<!-- ``` -->

<!-- Working with the Multivariate Normal -->
<!-- === -->
<!-- - Now let's simulate many multivariate normals.  -->
<!-- - Each row is a different sample from this multivariate normal distribution. -->
<!-- ```{r} -->
<!-- rmvnorm(n = 3, mean = rep(0, 2), sigma = diag(2)) -->
<!-- ``` -->




<!-- Load in data -->
<!-- === -->
<!-- ```{r} -->
<!-- # read in data -->
<!-- Y <- structure(c(59, 43, 34, 32, 42, 38, 55, 67, 64,  -->
<!--                  45, 49, 72, 34, 70, 34, 50, 41, 52,  -->
<!--                  60, 34, 28, 35, 77, 39, 46, 26, 38,  -->
<!--                  43, 68, 86, 77, 60, 50, 59, 38, 48,  -->
<!--                  55, 58, 54, 60, 75, 47, 48, 33),  -->
<!--                .Dim = c(22L, 2L), .Dimnames = list(NULL,  -->
<!--                 c("pretest", "posttest"))) -->
<!-- # number of observations -->
<!-- ``` -->

<!-- Quick calculations -->
<!-- ```{r} -->
<!-- (n <- dim(Y)[1]) -->
<!-- ybar <- apply(Y,2,mean) -->
<!-- Sigma <- cov(Y) -->
<!-- ``` -->




Detailed Takeaways on Background 
===

- Understanding vectors, matrices and notation  
- Understanding how to write multivariate notation for a conceptual problem
- Understanding how to write general multivariate notation
- Background on linear algebra 
- Determinants, traces, quadratic forms
- Knowing how to do simple proofs such as the exercises from class 

Detailed Takeaways on Multivariate Normal Models
===

- The multivariate normal distribution (MVN)
- Exercise with the MVN


Proof of Lemma 1
===

$$tr(AB) = tr(BA)$$

Proof: Suppose that $A_{n \times n}$ and $B_{n \times n}.$

Recall that by definition $tr(A) = \sum_{i=1}^n a_{ii}.$ By definition
\begin{align}
tr(AB) &= \sum_{i=1}^n (AB)_{ii} \\
& \sum_{i=1}^n \sum_{j=1}^n a_{ij} b_{ji}\\
& \sum_{i=1}^n \sum_{j=1}^n b_{ji} a_{ij} \\
&= \sum_{i=1}^n (BA)_{ii} 
= tr(BA)
\end{align}

