
---
title: "Module XX: Linear Regression"
author: "Rebecca C. Steorts"
date: Based Upon Hoff, Chapter 9
output: 
     beamer_presentation:
      includes: 
          in_header: custom2.tex
font-size: 8px
---



Agenda
===

- Motivation: oxygen uptake example 
- Linear regression
- Multiple and Multivariate Linear Regression
- Background on the Euclidean norm and argmin 
- Ordinary Least Squares + Exercises 


Oxygen uptake case study
===

Experimental design: 12 male volunteers.

\begin{enumerate}
\item 6 men take part in a randomized aerobics program
\item 6 remaining men take part in a randomized running program
\item \textcolor{blue}{The maximal $O_2$ uptake measured before and after the 12 week program} 
\item \textcolor{blue}{The change in maximal $O_2$ uptake is then calculated for each individual}
\end{enumerate}

What type of exercise is the most beneficial?

\vspace*{2em}

Full details of the study can be found in Hoff, page 149-151. 

Data
===
```{r}
# 0 is running
# 1 is aerobic exercise
x1<-c(0,0,0,0,0,0,1,1,1,1,1,1)
# x2 is age
x2<-c(23,22,22,25,27,20,31,23,27,28,22,24)
# change in maximal oxygen uptake
y<-c(-0.87,-10.74,-3.27,-1.97,7.50,
     -7.25,17.05,4.96,10.40,11.05,0.26,2.51)
```

Exploratory Data Analysis
===
```{r, echo=FALSE}
par(mfrow=c(1,1))
plot(y~x2,pch=16,xlab="age",ylab="change in maximal oxygen uptake", 
     col=c("red","green")[x1+1])
legend(27,0,legend=c("aerobic","running"),pch=c(16,16),col=c("green","red"))
```

Data analysis
===

$y$ = change in maximal oxygen uptake (scalar)

$x_1$ = exercise indicator (0 for running, 1 for aerobic)

$x_2$ = age

How can we estimate $p(y \mid x_1, x_2)?$

Linear regression
===
Assume that smoothness is a function of age.

For each group,

$$y = \beta_o + \beta_1 x_2 + \epsilon$$

Linearity means **linear in the parameters** ($\beta$'s). 

Linear regression
===

We could also try the model

$$y = \beta_o + \beta_1 x_2 +  \beta_2 x_2^2 + \beta_3 x_2^3 + \epsilon$$

which is also a linear regression model. 

Notation 
===
- $X_{n\times p}$: regression features or covariates (design matrix)
- $\bx_i$: $i$th row vector of the regression covariates
- $\by_{n\times 1}$: response variable (vector)
- $\bbeta_{p \times 1}$: vector of regression coefficients 


Notation (continued)
===
$$\bm{X}_{n \times p} = 
\left( \begin{array}{cccc}
x_{11} & x_{12} & \ldots&  x_{1p}\\
x_{21} & x_{22} & \ldots& x_{2p} \\
x_{i1} & x_{i2} & \ldots& x_{ip} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n1} & x_{n2} &\ldots& x_{np}
\end{array} \right).
$$

- A column of x represents a particular covariate we might be interested in, such as age of a person. 

- Denote $x_i$ as the ith \textcolor{red}{row vector} of the $X_{n \times p}$ matrix. 

\[  x_{i}= \left( \begin{array}{c}
x_{i1}\\
\textcolor{red}{x_{i2}}\\
\vdots\\
x_{ip}
\end{array} \right) \]

Notation (continued)
===
\[  \bbeta= \left( \begin{array}{c}
\beta_1\\
\beta_2\\
\vdots\\
\beta_p
\end{array} \right) \]

\[  \by= \left( \begin{array}{c}
y_1\\
y_2\\
\vdots\\
y_n
\end{array} \right) \]

$$\by_{n \times 1} = 
X_{n \times p} \bbeta_{p \times 1} + \bm{\epsilon}_{n \times 1}$$





Regression models
===
How does an outcome $\bm{y}$ vary as a function of the covariates which we represent as $X_{n\times p}$ matrix?


- Can we predict $\bm{y}$ as a function of each row in the matrix $X_{n\times p}$ denoted by $\bx_i.$
- Which $\bx_i$'s have an effect?

Such questions can be assessed via a linear regression model $p(\by \mid X).$


Multiple linear regression
===
Consider the following: 

$$y_i = \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} + 
\beta_4 x_{i4} + \epsilon_i$$

where
\begin{align}
x_{i1} &= 1 \; \text{for subject} \; i \\
x_{i2} &= 0 \; \text{for running}; \text{1 for aerobics}  \\
x_{i3} &= \text{age of subject i}\\
x_{i4} &= x_{i2} \times x_{i3} 
\end{align}

Under this model,
$$E[\bm{y} \mid \bm{x}] = \beta_1 + \beta_3 \times age \; \text{if} \; x_2=0$$
$$E[\bm{y} \mid \bm{x}] = (\beta_1 + \beta_2) + (\beta_3 + \beta_4)\times age \; \text{if} \; x_2=1 $$



Least squares regression lines 
===
```{r, echo=FALSE}
par(mfrow=c(2,2),mar=c(3,3,1,1),mgp=c(1.75,.75,0))

plot(y~x2,pch=16,col=c("red","green")[x1+1],ylab="change in maximal oxygen uptake",xlab="",xaxt="n")
abline(h=mean(y[x1==0]),col="red") 
abline(h=mean(y[x1==1]),col="green")
mtext(side=3,expression(paste(beta[3]==0,"  ",beta[4]==0)) )

plot(y~x2,pch=16,col=c("red","green")[x1+1],xlab="",ylab="",xaxt="n",yaxt="n")
abline(lm(y~x2),col="red")
abline(lm((y+.5)~x2),col="green")
mtext(side=3,expression(paste(beta[2]==0,"  ",beta[4]==0)) )

plot(y~x2,pch=16,col=c("red","green")[x1+1],
     xlab="age",ylab="change in maximal oxygen uptake" )
fit<-lm(y~x1+x2)
abline(a=fit$coef[1],b=fit$coef[3],col="red")
abline(a=fit$coef[1]+fit$coef[2],b=fit$coef[3],col="green")
mtext(side=3,expression(beta[4]==0)) 

plot(y~x2,pch=16,col=c("red","green")[x1+1],
     xlab="age",ylab="",yaxt="n")
abline(lm(y[x1==0]~x2[x1==0]),col="red")
abline(lm(y[x1==1]~x2[x1==1]),col="green")
```

Multivariate Setup 
===
Let's assume that we have data points $(x_i, y_i)$ available for all  $i=1,\ldots,n.$

- $y$ is the response variable
\[  \by= \left( \begin{array}{c}
y_1\\
y_2\\
\vdots\\
y_n
\end{array} \right)_{n \times 1} \]

- $\bx_{i}$ is the $i$th row of the design matrix $X_{n \times p}.$

Consider the regression coefficients

\[  \bbeta = \left( \begin{array}{c}
\beta_{1}\\
\beta_{2}\\
\vdots\\
\beta_{p}
\end{array} \right)_{p \times 1} \]

Normal Regression Model
===
The Normal regression model specifies that

- $E[Y\mid \bx_i]$ is linear and
- the sampling variability around the mean is independently and identically (iid) drawn from a normal distribution

\begin{align}
Y_i &= \bbeta^T \bx_i + \bm{\epsilon}_i\\
\epsilon_1,\ldots,\epsilon_n &\stackrel{iid}{\sim} \text{Normal}(0,\sigma^2)
\end{align}

This implies $Y_i \mid \bbeta, \bx_i \sim \text{Normal}(\bbeta^T \bx_i,\sigma^2).$




Multivariate Bayesian Normal Regression Model
===

We can re-write this as a multivariate regression model as:

$$\by \mid X,\bbeta, \sigma^2 \sim \text{MVN}( X\bbeta, \sigma^2 I_p).$$
 
We can specify a multivariate Bayesian model as: 

$$\by \mid X,\bbeta, \sigma^2 \sim \text{MVN}( X\bbeta, \sigma^2 I_p)$$
$$\bbeta \sim \text{MVN}(0, \tau^2 I_p),$$

where $\sigma^2, \tau^2$ are known.

Likelihood
===

The likelihood is 
\begin{align}
&p(y_1,\ldots,y_n \mid x_1,\ldots x_n, \bbeta, \sigma^2) \\
&= \prod_{i=1}^n p(\by_i \mid \bx_i, \bbeta, \sigma^2) \\
&(2\pi \sigma^2 )^{-n/2} \exp\{
\frac{-1}{2\sigma^2} \sum_{i=1}^n (\by_i - \bbeta^T \bx_i)^2
\}\\
&= (2\pi \sigma^2 )^{-n/2} \exp\{\textcolor{black}{-\frac{1}{2}} (\by - X\bbeta)^T\textcolor{black}{(\sigma^2)^{-1} I_p}(\by - X\bbeta)\}
\end{align}

Background
===

The Euclidean norm ($L^2$ norm or square root of the sum of squares) of $\boldsymbol{y} = (y_1, \ldots, y_n)$ is defined by 

$$ \|\boldsymbol{y}\|_2 = \sqrt{y_1^2 + \ldots + y_n^2}.$$
It follows that 

$$ \|\boldsymbol{y}\|_2^2 = y_1^2 + \ldots + y_n^2.$$
\vspace*{1em}

\textbf{Why do we use this notation?} It's compact and convenient! 

Background
===

We would like to find $$\argmin_{\bbeta \in \R^p} \|\by-X\bbeta\|_2^2,$$

where the $\argmin$ (the arguments of the minima) are the points or elements of the domains of some function as which the functions values are minimized. 


Ordinary Least Squares
===
We can estimate the coefficients $\hat{\bm{\beta}} \in \R^p$ by least squares:
$$\hat{\bm{\beta}} = \argmin_{\bbeta \in \R^p} \|\by-X\bbeta\|_2^2$$


One can show that 
$$\hat{\bm{\beta}} = (X^T X)^{-1} X^T \by$$


\bigskip
The fitted values are
$$\textcolor{black}{\hat{\bm{y}}} = X\hat{\bm{\beta}} = X(X^T X)^{-1} X^T \by$$
This is a linear function of $\by$, $\hy = H\by$, where
$H=X(X^T X)^{-1} X^T$ is sometimes called the **hat matrix**.

Exercise 1 (OLS)
===
Let SSR denote sum of squared residuals. 
$$ \min_{\bbeta} SSR(\bbeta) = \min_{\bbeta} \|\by-X{\bm{\beta}}\|_2^2$$
Show that  $$\hat{\bm{\beta}}  = (X^TX)^{-1}X^T\by.$$

Ordinary Least squares estimation
===
Proof: Observe
\begin{align}
 \frac{\partial SSR(\bbeta)}{\partial \bbeta} 
 &:=  \frac{\partial \|\by-X{\bm{\beta}}\|_2^2}{\partial \bbeta} \\
&= \frac{\partial (\by-X\bbeta)^T(\by-X\bbeta)}{\partial \bbeta} \\
&= \frac{\partial \by^T\by - 2\bbeta^TX^T\by + \hbeta^T(X^TX)\bbeta}{\partial \bbeta}\\
& = -  2X^T\by + 2X^TX\bbeta
\end{align}

This implies $-X^T\by + X^TX\bbeta= 0 \implies \hat{\bm{\beta}}_{ols}  = (X^TX)^{-1}X^T\by.$
\vskip 1em

This is called the **ordinary least squares estimator**. How do we know it is unique?

Exercise 2 (OLS)
===
Show that 

$$\hat{\bm{\beta}} \sim MVN(\bbeta, \sigma^2  (X^TX)^{-1}).$$

Ordinary Least squares estimation
===
Proof: Recall

$$\hbeta  = (X^TX)^{-1}X^T\bY. $$
\vskip 1em

$$E(\hat{\bm{\beta}} ) = E[(X^TX)^{-1}X^T\bY]= 
(X^TX)^{-1}X^T E[\bY] = (X^TX)^{-1}X^TX
\bbeta.$$

\vskip 1em
\begin{align}
\Var(\hat{\bm{\beta}}) &= \Var\{ (X^TX)^{-1}X^T\bY\}\\
 &=
(X^TX)^{-1}X^T \sigma^2 I_n X (X^TX)^{-1}\\
& = \sigma^2  (X^TX)^{-1}
\end{align}

$$\hat{\bm{\beta}} \sim MVN(\bbeta, \sigma^2  (X^TX)^{-1}).$$

Recall data set up
===
\footnotesize
```{r}
# running is 0, 1 is aerobic
x1<-c(0,0,0,0,0,0,1,1,1,1,1,1)
# age
x2<-c(23,22,22,25,27,20,31,23,27,28,22,24)
# change in maximal oxygen uptake
y<-c(-0.87,-10.74,-3.27,-1.97,7.50,
     -7.25,17.05,4.96,10.40,11.05,0.26,2.51)
```
Recall data set up
===
\footnotesize
```{r}
(x3 <- x2) #age
(x2 <- x1) #aerobic versus running 
(x1<- seq(1:length(x2))) #index of person
(x4 <- x2*x3)
```

Recall data set up
===
\footnotesize
```{r}
(X <- cbind(x1,x2,x3,x4))
```

OLS estimation in R
===
\footnotesize
```{r}
## using the lm function
fit.ols<-lm(y~ X[,2] + X[,3] +X[,4])
summary(fit.ols)$coef
```


